Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-14 talk/iwtc11/paper.tex
typo
spellchecker

Loop peeling is achieved by appending a copy of the traced iteration at
the end of the loop. The copy  is inlined to make the two parts form a
consitant two iteration trace. 
The first part (called preamble) finishes with the jump the the second part
(called peeled loop). The second part ends up with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every other iteration.
The trace from Figure~\ref{fig:unopt-trace} would after this operation become
the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
preamble while line 15-27 shows the peeled loop.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the pereamble, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

XXX gcc calles this store-sinking and I'm sure there are some
references in the literature (none at hand though). This is a ``typical''
compiler optimization.

@@

\subsection{Pure operation reusage}
If a pure operation appears more than once in the trace with same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. When that is combined with loop
peeling, the single execution of the operation is placed in the
preamble. That is, loop invariant pure operations are moved out of the
loop. The interactions here are the same as in the previous
section. That is, a vector, $H$, of variables produced in the preamble
and used in the peeled loop needs to be constructed. Then the jump and
input arguments are updated according to
Equation~\ref{eq:heap-inputargs} and Equation~\ref{eq:heap-jumpargs}.

@@

\subsection{Conclutions}
In this paper we have studied loop invariant code motion during trace
compilation. We claim that loop peeling is a very convenient solution
here since it fits well with other trace optimizations. By peeling of
the first iteration and optimizing the resulting two iteration trace
as a single trace, several standard optimizations can be 
used unchanged. The only interaction needed between the loop peeling
and the other 
optimizations is during the constructing of the jump arguments
connecting the peeled of iteration (the preamble) with the loop body. This
approach 
turns standard optimizations such as redundant guard removal, heap
caching, pure operation reusage and allocation removals into loop
invariant code motion optimizations.

>>>>

Loop peeling is achieved by appending a copy of the traced iteration at
the end of the loop. The copy  is inlined to make the two parts form a
consistent two iteration trace. 
The first part (called preamble) finishes with the jump the the second part
(called peeled loop). The second part ends up with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every other iteration.
The trace from Figure~\ref{fig:unopt-trace} would after this operation become
the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
preamble while line 15-27 shows the peeled loop.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the preamble, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

XXX gcc calls this store-sinking and I'm sure there are some
references in the literature (none at hand though). This is a ``typical''
compiler optimization.

@@

\subsection{Pure operation reuse}
If a pure operation appears more than once in the trace with same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. When that is combined with loop
peeling, the single execution of the operation is placed in the
preamble. That is, loop invariant pure operations are moved out of the
loop. The interactions here are the same as in the previous
section. That is, a vector, $H$, of variables produced in the preamble
and used in the peeled loop needs to be constructed. Then the jump and
input arguments are updated according to
Equation~\ref{eq:heap-inputargs} and Equation~\ref{eq:heap-jumpargs}.

@@

\subsection{Conclusions}
In this paper we have studied loop invariant code motion during trace
compilation. We claim that loop peeling is a very convenient solution
here since it fits well with other trace optimizations. By peeling of
the first iteration and optimizing the resulting two iteration trace
as a single trace, several standard optimizations can be 
used unchanged. The only interaction needed between the loop peeling
and the other 
optimizations is during the constructing of the jump arguments
connecting the peeled of iteration (the preamble) with the loop body. This
approach 
turns standard optimizations such as redundant guard removal, heap
caching, pure operation reuse and allocation removals into loop
invariant code motion optimizations.
